Characterizing Stellar Streams by ML

### 2023-08-31 src/tex/ms.tex
[showyourwork] overleaf sync

\section{Introduction} \label{sec:intro}
    %
    Stellar streams are the disrupted
    remnants of globular clusters and satellite galaxies, and their census has
    grown extensively in the Milky Way thanks to all-sky astrometric missions
    like {\it Gaia} and low-surface brightness photometric surveys (e.g., \citealt{2018ApJ...862..114S, 2021ApJ...914..123I}).
    The formation of streams can be traced back to a progenitor cluster, whose stars are gradually lost to the
    tidal field of the host galaxy and extend along a series of similar orbits
    producing tidal tails, which can sometimes span several tens of degrees across
    the sky. The orbital proximity of stars belonging to a stream makes them
    sensitive tracers of the mass distribution of galaxies, sourced by both the
    baryonic and dark matter components.

@@

            Let $\cmp{S,m}{\mbs{\theta}}(\phi_1)$ represent the feed-forward
            neural network which outputs a vector of parameters characterizing
            the stream's photometric distribution at a given $\phi_1$. The
            network has 2 outputs -- $\mu, \sigma_\mu$ -- as we only consider
            the distance track of the isochrone, holding fixed stellar
            population parameters internal to modeling the isochrone, such as
            its age and metallicity.  In principle we might include parameters
            of the isochrone in the model, now $\mcal{I}(\mbs{\theta}, \gamma)$,
            where $\theta = {a, Z, ...}$ and any other generating parameters of
            the isochrone.
            % FIXME! DO this Likewise, the dust map may be treated as a prior and it's errors incorporated into the data errors.
            However, stellar streams are generally poorly fit by theoretic
            isochrones, at least ones with realistic ages and metallicities.
            Consequently including the age and metallicity as model parameters
            does not meaningfully contribute to constraining results on the
            stream's stellar population. Moreover, age and metallicity are
            partially degenerate with the distance modulus, working counter to
            the goal of fitting a distance track to the stream. An alternate
            approach is to use data-driven isochrones instead of theoretic ones.
            the age and metallicity are not meaningful quantities. In others,
            e.g. with a suite of models or a trained emulator, the data-driven
            isochrones might have realistic ages and metallicities and thus
            meaningfully contribute to the model.
            % Is this true? re-read Uniform Modeling of 13 streams
            We defer this to a future work.

@@

            % TODO: rephrase
            By modeling jointly the astrometrics and photometrics we may tie
            together components of the models.  In particular, the distance
            modulus in the photometric model and the parallax in the astrometric
            model are both measures of the distance track of the stream.  We
            introduce a prior to connect these components of the model:
            \begin{align}
                \prior(\{\mu_I, \mu_\parallax \} | \phi_1) &= \delta( \rm{dist}(\mu_I) - \rm{dist}(\parallax) )(\phi_1), \label{eq:distance_track_prior}
                % \prior(\{\sigma_I, \sigma_\parallax \}(\phi_1)) &= \delta( \rm{dist}(\mu_I) - \rm{dist}(\parallax) )(\phi_1) \label{eq:distance_width_prior}
            \end{align}
            % TODO: should we show that these are different distance functions? or is it obvious
            where $\delta$ is the Dirac delta function
            \citep{Dirac1947}, and enforces that the distance modulus
            track and parallax track must be equal (when converted to the actual
            distance) as a function of $\phi_1$.  The widths are similarly
            constrained. The distance and first-order error conversion are given
            by
            \begin{align}
                \mu_I(\parallax) &= 10 - 5\log_{10}(\parallax) \\
                \sigma_I(\sigma_\parallax)  &\approxeq \frac{5}{\ln{10}} \mu_I \lvert\sigma_\parallax\rvert.
            \end{align}

@@

        \begin{figure}
            \script{mock/plot/photometric_background_selection.py}
            \centering
            \includegraphics[width=1\linewidth]{figures/mock/photometric_background_selection.pdf}
            \caption{
                % 
                \textbf{Top panel:} Astrometric coordinates ($\phi_1$, $\phi_2$) for
                the mock data set.  Light blue points are the stream and background
                within an ``on"-stream region. Black points are in the ``off"-stream
                region used to train the background photometric model.
                % 
                \textbf{Bottom panel:} Astrometric coordinates ($\phi_1$, $\phi_2$)
                for the mock data set.  Black points are in the ``off"-stream region
                from the region defined in the astrometric coordinates.  Light blue
                points are the stream and background from the on-stream region. The
                stream's isochrone is apparent as an overdensity around (g-r,
                g)$\approx (0, 20)$.  }
            \label{fig:mock_data_photometric_background_selection}
        \end{figure}

@@

        We find that the stream is successfully recovered in both astrometric and
        photometric coordinates, with $\variable{output/mock/ncorrect_variable.txt}\%$ of stream
        stars being recovered with membership probability greater than $80\%$.
        In the bottom two rows of \autoref{fig:mock_data_result} we illustrate the
        synthetic data color-coded by stream membership probability (top) and
        background membership probability (background) for different $\phi_1$
        slices. Importantly, the photometric portion of the mixture model
        successfully identifies the stream in magnitude space, showing a high stream
        membership probability where the stream is present, and a low probability
        ($\sim 0$) where the stream is absent.
        The false positive rate (i.e., stars misidentified as stream members) for this test
        is found to be $\variable{output/mock/falseident_variable.txt}\%$ (3 stars) when cutting on
        the 80\% membership probability threshold.
        % The perfect recovery rate and low false identification rate demonstrate the efficacy
        % of this method.

@@

% \section{Theory of Mixture Density Networks (MDNs)} \label{app:theory_of_mixture_density_networks}

%     Mixture models (MMs) are a statistical method to represent a population as a set of a sub-populations specified by probability distributions. These distributions are characterized by their parameters and mixture by a corresponding set of mixture weights. Mixture Density Networks (MDNs) 
%     are a machine learning method to combine a MM with neural networks, increasing the flexibility of the model. An MDN replaces the scalar distribution parameters of the MM with neural networks, allowing the distribution parameters and mixture weights to vary as smooth functions \citep{Bishop1994}. Functional parameters allows the model to capture variation over those parameters that is not possible with scalar-valued models. Moreover, by using neural networks the inferred functional forms are ``model-free" and driven entirely by the data.

%     % Quote from https://mikedusenberry.com/mixture-density-networks
%     % A mixture density network (MDN) is an interesting model formalism built within the general framework of neural networks and probability theory for working on supervised learning problems in which the target variable cannot be easily approximated by a single standard probability distribution. Essentially an MDN allows one to model a conditional probability distribution as a mixture of distributions, in which the individual distributions and the corresponding mixture coefficients are parameterized by functions of the inputs

%     For the MDN we take $x_{\cdot, 0} \equiv x_{\cdot}$ (indexed from 0), which is the first feature column of $\mbf{X}$, as the `independent' coordinate over which the MDN parameters are conditioned. In the context of streams $x_{\cdot, 0} \equiv \phi_1$ is the longitude coordinate in a reference frame rotated such that the stream is aligned along $\phi_1$.

@@

%     \begin{equation} \label{eq:general_mixture_network}
%         \pdf(\nth{\mbs{x}} | \mbs{\theta}(\phi_1))
%         = \sum_{\mrm{q} \in \mbs{I_Q}} \ \qth{f}(\phi_1) \qth{P}(\nth{x}|\qth{\mbs{\theta}}(\phi_1),q\!=\!\mrm{q})
%     \end{equation}
%     The mixture coefficients $\qth{f}$ are normalized s.t.
%     \begin{equation}
%         \sum_{\mrm{q}} \qth{f}(\phi_1) = 1 \qquad \forall \phi_1
%     \end{equation}
%     In practice this is enforced by defining
%     \begin{equation}
%         f_{b}(\phi_1) = 1 - \sum_{\rm{q}\in \mbs{I_Q}\backslash b} \qth{f}(\phi_1)
%     \end{equation}
%     for some ``background" index $b \in \mbs{I_Q}$.

%     Assuming the measurement of each datum is independent, the likelihood is a sum over all data:
%     \begin{equation}
%         \mcal{L}\left(\mbf{x}_n | \mbs{\theta}(\phi_1)\right) = \sum_{q\in\mbs{I_Q}} \qth{P}\left(\nth{\mbs{x}} | \qth{\mbs{\theta}}(\phi_1)\right).
%     \end{equation}

@@

%     \begin{equation}
%         \ln\posterior\left(\mbf{X} | \mbs{\theta}(\mbs{x})\right) = \ln\mcal{L}\left(\mbf{X} | \mbs{\theta}(\mbs{x})\right) +  \sum_{n \in \mbs{I_N}} \ln \left[ \sum_{q\in\mbs{I_Q}} \qth{\prior}\left(\qth{\mbs{\theta}}(\mbf{x}_n)\right) \right].
%     \end{equation}

@@

% \section{Plots} \label{app:developing_plots}

%     \begin{figure*}
%         \centering
%         \includegraphics[scale=0.7]{figures-not-syw/gd1/tot_stream_and_spur_prob.png}
%         \caption{Caption}
%         \label{fig:gd1-astrometrics}
%     \end{figure*}

%     \begin{figure*}
%         \centering
%         \includegraphics[scale=0.8]{figures-not-syw/gd1/Unknown.png}
%         \caption{Caption}
%         \label{fig:gd1-distance_modulus}
%     \end{figure*}

%     \begin{figure}
%         \centering
%         \includegraphics[scale=0.6]{figures-not-syw/gd1/Unknown-2.png}
%         \caption{Caption}
%         \label{fig:gd1-cmd}
%     \end{figure}

%     \begin{figure*}
%         \centering
%         \includegraphics[scale=0.6]{figures-not-syw/gd1/Unknown-4.png}
%         \caption{Caption}
%         \label{fig:gd1-cmds}
%     \end{figure*}

>>>>

\section{Introduction} \label{sec:intro}

    Stellar streams are the disrupted remnants of globular clusters and
    satellite galaxies, and their census has grown extensively in the Milky Way
    thanks to all-sky astrometric missions like {\it Gaia} and low-surface
    brightness photometric surveys (e.g., \citealt{2018ApJ...862..114S,
    2021ApJ...914..123I}).  The formation of streams can be traced back to a
    progenitor cluster, whose stars are gradually lost to the tidal field of the
    host galaxy and extend along a series of similar orbits producing tidal
    tails, which can sometimes span several tens of degrees across the sky. The
    orbital proximity of stars belonging to a stream makes them sensitive
    tracers of the mass distribution of galaxies, sourced by both the baryonic
    and dark matter components.

@@

            Let $\cmp{S,m}{\mbs{\theta}}(\phi_1)$ represent the feed-forward
            neural network which outputs a vector of parameters characterizing
            the stream's photometric distribution at a given $\phi_1$. The
            network has 2 outputs -- $\mu, \sigma_\mu$ -- as we only consider
            the distance track of the isochrone, holding fixed stellar
            population parameters internal to modeling the isochrone, such as
            its age and metallicity.  In principle we might include parameters
            of the isochrone in the model, now $\mcal{I}(\mbs{\theta}, \gamma)$,
            where $\theta = {a, Z, ...}$ and any other generating parameters of
            the isochrone.
            % FIXME! DO this Likewise, the dust map may be treated as a prior
            % and it's errors incorporated into the data errors.
            However, stellar streams are generally poorly fit by theoretic
            isochrones, at least ones with realistic ages and metallicities.
            Consequently including the age and metallicity as model parameters
            does not meaningfully contribute to constraining results on the
            stream's stellar population. Moreover, age and metallicity are
            partially degenerate with the distance modulus, working counter to
            the goal of fitting a distance track to the stream. An alternate
            approach is to use data-driven isochrones instead of theoretic ones.
            the age and metallicity are not meaningful quantities. In others,
            e.g. with a suite of models or a trained emulator, the data-driven
            isochrones might have realistic ages and metallicities and thus
            meaningfully contribute to the model.
            % Is this true? re-read Uniform Modeling of 13 streams
            We defer this to a future work.

@@

            % TODO: rephrase
            By modeling jointly the astrometrics and photometrics we may tie
            together components of the models.  In particular, the distance
            modulus in the photometric model and the parallax in the astrometric
            model are both measures of the distance track of the stream.  We
            introduce a prior to connect these components of the model:
            \begin{align}
                \prior(\{\mu_I, \mu_\parallax \} | \phi_1) &= \delta( \rm{dist}(\mu_I) - \rm{dist}(\parallax) )(\phi_1), \label{eq:distance_track_prior}
                % \prior(\{\sigma_I, \sigma_\parallax \}(\phi_1)) &= \delta( \rm{dist}(\mu_I) - \rm{dist}(\parallax) )(\phi_1) \label{eq:distance_width_prior}
            \end{align}
            % TODO: should we show that these are different distance functions?
            % or is it obvious
            where $\delta$ is the Dirac delta function \citep{Dirac1947}, and
            enforces that the distance modulus track and parallax track must be
            equal (when converted to the actual distance) as a function of
            $\phi_1$.  The widths are similarly constrained. The distance and
            first-order error conversion are given by
            \begin{align}
                \mu_I(\parallax) &= 10 - 5\log_{10}(\parallax) \\
                \sigma_I(\sigma_\parallax)  &\approxeq \frac{5}{\ln{10}} \mu_I \lvert\sigma_\parallax\rvert.
            \end{align}

@@

        \begin{figure}
            \script{mock/plot/photometric_background_selection.py}
            \centering
            \includegraphics[width=1\linewidth]{figures/mock/photometric_background_selection.pdf}
            \caption{
                %
                \textbf{Top panel:} Astrometric coordinates ($\phi_1$, $\phi_2$)
                for the mock data set.  Light blue points are the stream and
                background within an ``on"-stream region. Black points are in
                the ``off"-stream region used to train the background
                photometric model.
                %
                \textbf{Bottom panel:} Astrometric coordinates ($\phi_1$,
                $\phi_2$) for the mock data set.  Black points are in the
                ``off"-stream region from the region defined in the astrometric
                coordinates.  Light blue points are the stream and background
                from the on-stream region. The stream's isochrone is apparent as
                an overdensity around (g-r, g)$\approx (0, 20)$.
                %
            }
            \label{fig:mock_data_photometric_background_selection}
        \end{figure}

@@

        We find that the stream is successfully recovered in both astrometric
        and photometric coordinates, with
        $\variable{output/mock/ncorrect_variable.txt}\%$ of stream stars being
        recovered with membership probability greater than $80\%$.  In the
        bottom two rows of \autoref{fig:mock_data_result} we illustrate the
        synthetic data color-coded by stream membership probability (top) and
        background membership probability (background) for different $\phi_1$
        slices. Importantly, the photometric portion of the mixture model
        successfully identifies the stream in magnitude space, showing a high
        stream membership probability where the stream is present, and a low
        probability ($\sim 0$) where the stream is absent.  The false positive
        rate (i.e., stars misidentified as stream members) for this test is
        found to be $\variable{output/mock/falseident_variable.txt}\%$ (3 stars)
        when cutting on the 80\% membership probability threshold.
        % The perfect recovery rate and low false identification rate
        % demonstrate the efficacy of this method.

@@

% \section{Theory of Mixture Density Networks (MDNs)}
% \label{app:theory_of_mixture_density_networks}

%     Mixture models (MMs) are a statistical method to represent a population as
%     a set of a sub-populations specified by probability distributions. These
%     distributions are characterized by their parameters and mixture by a
%     corresponding set of mixture weights. Mixture Density Networks (MDNs) are
%     a machine learning method to combine a MM with neural networks, increasing
%     the flexibility of the model. An MDN replaces the scalar distribution
%     parameters of the MM with neural networks, allowing the distribution
%     parameters and mixture weights to vary as smooth functions
%     \citep{Bishop1994}. Functional parameters allows the model to capture
%     variation over those parameters that is not possible with scalar-valued
%     models. Moreover, by using neural networks the inferred functional forms
%     are ``model-free" and driven entirely by the data.

%     % Quote from https://mikedusenberry.com/mixture-density-networks % A
%     mixture density network (MDN) is an interesting model formalism built
%     within the general framework of neural networks and probability theory for
%     working on supervised learning problems in which the target variable
%     cannot be easily approximated by a single standard probability
%     distribution. Essentially an MDN allows one to model a conditional
%     probability distribution as a mixture of distributions, in which the
%     individual distributions and the corresponding mixture coefficients are
%     parameterized by functions of the inputs

%     For the MDN we take $x_{\cdot, 0} \equiv x_{\cdot}$ (indexed from 0),
%     which is the first feature column of $\mbf{X}$, as the `independent'
%     coordinate over which the MDN parameters are conditioned. In the context
%     of streams $x_{\cdot, 0} \equiv \phi_1$ is the longitude coordinate in a
%     reference frame rotated such that the stream is aligned along $\phi_1$.

@@

%     \begin{equation} \label{eq:general_mixture_network} \pdf(\nth{\mbs{x}} |
%         \mbs{\theta}(\phi_1)) = \sum_{\mrm{q} \in \mbs{I_Q}} \ \qth{f}(\phi_1)
%         \qth{P}(\nth{x}|\qth{\mbs{\theta}}(\phi_1),q\!=\!\mrm{q})
%         \end{equation} The mixture coefficients $\qth{f}$ are normalized s.t.
%         \begin{equation} \sum_{\mrm{q}} \qth{f}(\phi_1) = 1 \qquad \forall
%         \phi_1 \end{equation} In practice this is enforced by defining
%         \begin{equation} f_{b}(\phi_1) = 1 - \sum_{\rm{q}\in
%         \mbs{I_Q}\backslash b} \qth{f}(\phi_1) \end{equation} for some
%         ``background" index $b \in \mbs{I_Q}$.

%     Assuming the measurement of each datum is independent, the likelihood is a
%     sum over all data: \begin{equation} \mcal{L}\left(\mbf{x}_n |
%     \mbs{\theta}(\phi_1)\right) = \sum_{q\in\mbs{I_Q}}
%     \qth{P}\left(\nth{\mbs{x}} | \qth{\mbs{\theta}}(\phi_1)\right).
%     \end{equation}

@@

%     \begin{equation} \ln\posterior\left(\mbf{X} | \mbs{\theta}(\mbs{x})\right)
%         = \ln\mcal{L}\left(\mbf{X} | \mbs{\theta}(\mbs{x})\right) +  \sum_{n
%         \in \mbs{I_N}} \ln \left[ \sum_{q\in\mbs{I_Q}}
%         \qth{\prior}\left(\qth{\mbs{\theta}}(\mbf{x}_n)\right) \right].
%         \end{equation}

@@
